# Read data {#Read Data}

## Load packages

```{r}

source("./00_dependencies.R")


```


## Load data

```{r}

####-----  List samples
samples <- str_replace(str_sort(list.files("../data/counts_mx"), numeric = TRUE),
                       "_filtered_feature_bc_matrix",
                       "")
print("Samples: ")
samples

####----- Create a named vector with paths to samples
samples.dirs <- sapply(samples, function(i) {
  d10x <- file.path('../data/counts_mx',
                    paste0(i, '_filtered_feature_bc_matrix'))
  d10x
})
print("Sample directories:")
samples.dirs

####----- Read count matrices
gex_data <- Read10X(data.dir = samples.dirs,
                    unique.features = TRUE,
                    strip.suffix = TRUE)
str(gex_data)

####----- Create seurat object
gex <- CreateSeuratObject(gex_data, 
                          project = "pSS",
                          min.cells = 0,
                          min.features = 0,
                          names.field = 1)
table(gex@meta.data$orig.ident)
nrow(gex@meta.data)

####-----  rm tmp objects
rm(samples, samples.dirs, gex_data)

####-----  change orig.ident for P007 and P008; libraries were generated and sequenced twice
gex@meta.data$orig.ident <- gsub("P007[ab]", "P007", gex@meta.data$orig.ident)
gex@meta.data$orig.ident <- gsub("P008[ab]", "P008", gex@meta.data$orig.ident)

table(gex@meta.data$orig.ident)

####-----  add pheno info
pheno <- read.table("../suppl/pheno/pSS_pheno_210623.csv", sep = ";", header = TRUE)
colnames(pheno)[c(1, 3)] <- c("Sample_id", "orig.ident")
pheno$orig.ident <- gsub("P007[ab]", "P007", pheno$orig.ident)
pheno$orig.ident <- gsub("P008[ab]", "P008", pheno$orig.ident)
pheno <- unique(pheno[, c(1, 3, 5, 7:8, 11:12)])
head(pheno); dim(pheno)

meta.data.tmp <- gex@meta.data
meta.data.tmp$cell <- row.names(meta.data.tmp)
meta.data.tmp <- merge(meta.data.tmp, pheno, by = "orig.ident", all.x = TRUE)
row.names(meta.data.tmp) <- meta.data.tmp$cell
meta.data.tmp$cell <- NULL

gex <- AddMetaData(gex, meta.data.tmp); head(gex@meta.data); table(gex@meta.data$orig.ident)

Idents(object = gex) <- "orig.ident"

print("GEX_RAW size:")
print(object.size(gex), units = "GB")

gex


```


## save rds GEX1

```{r include = FALSE}

saveRDS(gex, file = "../results/GEX_RAW.rds")

#this is GEX1

```


## Print sessionInfo()

```{r}

sessionInfo()


```





You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
