---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Dimensionality reduction, clusterimg, and cluster annotation {#Dimred1}

## Load packages

```{r}

source("./00_dependencies.R")

```


## Load data

```{r}

gex <- readRDS("../results/GEX2_FILTERED.rds")
annot <- readRDS("../suppl/annot/refdata-gex-GRCh38-2020-A.genes.gtf.annot.rds")

```


# Normalize and get variable features
 
```{r}

gex <- NormalizeData(gex, scale.factor = 10000)
gex <- FindVariableFeatures(gex, selection.method = "vst", nfeatures = 4000)

p1 <- LabelPoints(plot = VariableFeaturePlot(gex), 
                  points = head(VariableFeatures(gex), 80), 
                  repel = TRUE,
                  xnudge = 0,
                  ynudge = 0,
                  max.overlaps = Inf) + 
        theme_classic() +
        ggtitle(paste0(length(VariableFeatures(gex)), " Variable features - allCells (all features"))


#### ------ Subset VariableFeatures(gex) to only include biotype == "protein_coding"
VariableFeatures(gex) <-
  VariableFeatures(gex)[VariableFeatures(gex) %in%
                          annot$external_gene_name[annot$gene_biotype == "protein_coding"]]

p2 <- LabelPoints(plot = VariableFeaturePlot(gex), 
                  points = head(VariableFeatures(gex), 80), 
                  repel = TRUE,
                  xnudge = 0,
                  ynudge = 0,
                  max.overlaps = Inf) + 
        theme_classic() +
        ggtitle(paste0(length(VariableFeatures(gex)), " Variable features - allCells (protein_coding)"))

ggsave2("../results/VariableFeatures_allCells.png", 
        plot_grid(p1, p2, ncol = 2, labels = "AUTO", align = "h"), 
        width = 30, height = 15, unit = "cm")


```


## Scale data

```{r}

#### ----- Regressing out nFeature_RNA
gex <- ScaleData(gex, 
                 features = VariableFeatures(gex),
                 vars.to.regress = c("nFeature_RNA"))

```


## PCA

```{r}

#RunPCA() takes around 2h for this data set using 16 cores
gex <- RunPCA(
  gex,
  assay = "RNA",
  features = VariableFeatures(gex),
  npcs = 100,
  reduction.name = "pca_1",
  verbose = TRUE
)

#remove scale data slot to reduce Seurat object size
gex@assays$RNA@scale.data <- matrix(0)
gc()

```


## Harmony batch correction

```{r}

#RunHarmony() takes around 1h for this data set using 16 cores

gex <- RunHarmony(
  gex,
  group.by.vars = "orig.ident",
  reduction = "pca_1",
  reduction.save = "harmony_1",
  assay = "RNA",
  project.dim = FALSE, #project.dim = FALSE needed for seurat object v4.0.0??
  verbose = TRUE
)

```


# plot PCA and harmony

```{r}

p1 <-DimPlot(object = gex,
              reduction = "pca_1",
              pt.size = .1,
              group.by = "orig.ident",
              raster = FALSE) +
                  NoLegend() +
                  ggtitle("PCA_1, by Sample")

p2 <- DimPlot(object = gex,
              reduction = "harmony_1",
              pt.size = .1,
              group.by = "orig.ident",
              raster = FALSE) +
                  NoLegend() +
                  ggtitle("Harmony_1, by Sample")

ggsave2("../results/PCA_harmony_allCells_.png", 
        plot_grid(p1, p2, ncol = 2, labels = NULL, align = "h"), 
        width = 30, height = 15, unit = "cm")


```


# run UMAP

```{r}

gex <- RunUMAP(gex,
                dims = 1:50,
                reduction = "harmony_1",
                metric = "correlation",
                reduction.name = "umap_1",
                min.dist = 0.4, #local cell separation
                spread = .5, #global cell separation
                n.neighbors = 30, #30 on bianca!
                repulsion.strength = 0.4, #~2x min.dist, repulsion to cells fr annat cluster, global separation
                negative.sample.rate = 50, #ant ggr n.neighbors celler, global distance
                n.epochs = 100,
                n.components = 2
              )


```


## Clustering

```{r}

gex <- FindNeighbors(gex,
  reduction = "harmony_1",
  dims = 1:50,
  k.param = 30,
  verbose = TRUE
)

gex <- FindClusters(gex, 
                    resolution = 0.4, 
                    verbose = TRUE, 
                    graph.name = "RNA_nn")

table(gex@meta.data$seurat_clusters)

cat("GEX_DIMRED size: \n")
print(object.size(gex), units = "GB")

```


## hclust
### Check cluster relations

```{r}

cluster_means <-
  sapply(as.character(unique(gex$seurat_clusters)), function(x) {
    rowMeans(gex@assays$RNA@data[, gex$seurat_clusters == x])
  })

adj <- (1 - cor(cluster_means)) / 2
h <- hclust(as.dist(adj), method = "ward.D2")

png("../results/hclust_clusters_allCells_clustRes.png", 
    width = 1500, 
    height = 1500, 
    units = "px")
plot(as.dendrogram(h))
dev.off()


```


## Plot 2D UMAP for all cells

```{r}

p1 <- DimPlot(gex, 
                  reduction = "umap_1", 
                  pt.size = .1, 
                  group.by = "orig.ident", 
                  raster = FALSE) + 
                      NoLegend() + 
                      ggtitle(paste0(nrow(gex@meta.data), " cells, by sample"))  +
          ylab("UMAP 2") +
          xlab("UMAP 1") +
  theme(axis.text = element_blank(),
          axis.ticks = element_blank())

p2 <- DimPlot(gex, 
                  reduction = "umap_1", 
                  pt.size = .1, 
                  group.by = "seurat_clusters", 
                  label = TRUE, 
                  repel = TRUE, 
                  raster = FALSE) + 
                      NoLegend() + 
                      ggtitle(paste0("by cluster")) +
          ylab("UMAP 2") +
          xlab("UMAP 1") +
  theme(axis.text = element_blank(),
          axis.ticks = element_blank())

ggsave2("../results/UMAP_allCells_SFig2B.png", 
        plot_grid(p1, p2, ncol = 2, labels = NULL, align = "h"), 
        width = 30, height = 15, unit = "cm")


```


## Plot marker genes

```{r}

ig.features <- c("IGHA1", "IGHA2", "IGHG1", "IGHG2", "IGHG3", 
                 "IGHG4",  "IGHD", "IGHE", "IGHM")

b.markers <- c("CD79A", "CD79B", "MS4A1", "CD19", "CD27", 
               "IGHA1", "IGHD", "IGHM", "JCHAIN", "MME")

pbmc.markers <- c("MS4A1", "CD19", "CD27", "CD79A", "GNLY", 
                  "CD3E", "CD14", "LYZ", "CD8A")

my_pars <- c("nCount_RNA", "nFeature_RNA", "percent_mito", "percent_ribo", "percent_hb")

####-----  PBMC markers
p.pbmc <- FeaturePlot(gex, 
                        features = pbmc.markers, 
                        reduction = "umap_1", 
                        raster=FALSE, 
                        order = TRUE) 

ggsave2("../results/FeaturePlot_allCells_pbmcMarkers.png", 
        p.pbmc, 
        width = 25, height = 25, unit = "cm")

####-----  B markers
p.b.markers <- FeaturePlot(gex, 
                             features = b.markers, 
                             reduction = "umap_1", 
                             raster=FALSE, 
                             order = TRUE)
ggsave2("../results/FeaturePlot_allCells_BMarkers.png", 
        p.b.markers, 
        width = 30, height = 20, unit="cm")

####-----  IG markers
p.ig.markers <- FeaturePlot(gex, 
                              features = ig.features, 
                              reduction = "umap_1", 
                              raster = FALSE, 
                              order = TRUE)

ggsave2("../results/FeaturePlot_allCells_IGMarkers.png", 
        p.ig.markers, 
        width = 25, height = 25, unit = "cm")

####-----  my_pars
p.myPars <- FeaturePlot(gex, 
                          features = my_pars , 
                          reduction = "umap_1", 
                          raster = FALSE, 
                          order = TRUE)
ggsave2("../results/FeaturePlot_allCells_myPars.png", 
        p.myPars, 
        width = 25, height = 25, unit="cm")


```


#SFig2A

```{r}

p1 <- DimPlot(gex, 
                  reduction = "umap_1", 
                  pt.size = .05, 
                  group.by = "orig.ident", 
                  raster = TRUE) + 
                      NoLegend() + 
                      ggtitle(paste0(nrow(gex@meta.data), " cells, by sample"))  +
          ylab("UMAP 2") +
          xlab("UMAP 1") +
  theme(axis.text = element_blank(),
          axis.ticks = element_blank())

p2 <- DimPlot(gex, 
                  reduction = "umap_1", 
                  pt.size = .05, 
                  group.by = "seurat_clusters", 
                  label = TRUE, 
                  repel = TRUE, 
                  raster = TRUE) + 
                      NoLegend() + 
                      ggtitle(paste0("by cluster")) +
          ylab("UMAP 2") +
          xlab("UMAP 1") +
  theme(axis.text = element_blank(),
          axis.ticks = element_blank())


ig.features <- c("IGHA1", "IGHA2", "IGHG1", "IGHG2", "IGHG3", "IGHG4",  "IGHD", "IGHE", "IGHM")

pbmc.markers <- c("MS4A1", "CD19", "CD27", "CD79A", "GNLY", 
                  "CD3E", "CD14", "LYZ", "CD8A")

pig <- list()
for (i in ig.features) {
  
  pig[[i]] <- FeaturePlot(gex, 
                        reduction = "umap_1", 
                        features = i,
                        pt.size = 0.8) + 
    xlab("UMAP 1") + 
    ylab("UMAP 2") + 
    theme(legend.position = "none",
          axis.text = element_blank(),
          axis.ticks = element_blank())
  
}

pp <- list()
for (i in pbmc.markers) {
  
  pp[[i]] <- FeaturePlot(gex, 
                        reduction = "umap_1", 
                        features = i,
                        pt.size = 0.5) + 
    xlab("UMAP 1") + 
    ylab("UMAP 2") + 
    theme(legend.position = "none",
          axis.text = element_blank(),
          axis.ticks = element_blank())
  
}

p3 <- plot_grid(p1, p2, ncol = 1)
p4 <- plot_grid(plotlist = pig, ncol = 3)
p5 <- plot_grid(plotlist = pp, ncol = 3)

ggsave2(paste0("../results/SFig2B.png"), 
        plot_grid(p3, NULL, p4, NULL, p5, ncol = 5, rel_widths = c(1, 0.1, 1.5, 0.1, 1.5)) + bgcolor("white"), 
        width = 30, height = 15, unit = "cm")


```     


## Get cluster marker genes

```{r}

sample_size <- table(gex$seurat_clusters)
sample_size[sample_size > 100] <- 100

DGE_cells <- lapply(names(sample_size), function(x) {
  set.seed(1)
  sample(colnames(gex)[gex$seurat_clusters == x], size = sample_size[x])
})

DGE_cells <- unlist(DGE_cells)
DGE_temp<- gex[, DGE_cells]
DGE_temp <- SetIdent(DGE_temp, value = "seurat_clusters")
detable <- FindAllMarkers( DGE_temp, 
                           min.pct = 0.1, 
                           min.cells.feature = 0.1, 
                           max.cells.per.ident = 50, 
                           only.pos = TRUE, 
                           logfc.threshold = 0.1, 
                           assay = "RNA")

detable$pct.diff <- detable$pct.1 - detable$pct.2
write.csv(detable, 
          "../results/DGE_allCells_seurat_clusters.csv")

detable$log.pct.diff <- log2(detable$pct.1 / (detable$pct.2 + 0.01) )
detable <- detable[ detable$p_val < 0.05, ]

detable %>% group_by(cluster) %>% top_n(-30, p_val)  %>% top_n(20, pct.diff) -> top5
ord <-
  factor(sapply(unique(as.character(top5$gene)), function(x) {
    getcluster(DGE_temp,
               x,
               "seurat_clusters")
  }))

genes_to_plot <- unique(as.character(top5$gene))[order(as.numeric(as.character(ord)))]

pdf("../results/DGE_allCells_seurat_clusters_dotplot.pdf", width = 10, height = 50)
rafalib::mypar(1, 1, mar = c(6, 6, 1, 5))
plot_dots(gex, 
          genes_to_plot, 
          clustering = "seurat_clusters", 
          show_grid = T, 
          main = "top cluster markers", 
          cex.main = 1, font.main = 1, cex.col = 1.3, srt = 0, cex.row = 1.3)
dev.off()


```


## Manual cluster annotation

```{r}

annotation <- c(
  B_cell = 0,
  B_cell = 1,
  B_cell = 2,
  B_cell = 3,
  B_cell = 4,
  B_cell = 5,
  B_cell = 8,
  T_cell = 6,
  Monocyte = 7,
  Plasma_cell = 9
)

gex$celltype <- names(annotation)[match(gex$seurat_clusters, annotation)] 

p3 <- DimPlot(gex, 
                  reduction = "umap_1", 
                  pt.size = .01, 
                  group.by = "celltype", 
                  label = TRUE, 
                  repel = TRUE, 
                  raster = TRUE) + 
             ggtitle(paste0("cell types")) +
          ylab("UMAP 2") +
          xlab("UMAP 1") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")

ggsave2("../results/UMAP_allCells_celltypes_Fig1B.png", 
        plot_grid(p3, ncol = 1, labels = NULL, align = "h"), 
        width = 12, height = 12, unit = "cm")

counts <- as.data.frame.matrix(table(gex@meta.data$celltype, gex@meta.data$orig.ident))
write.csv(counts, "../results/cellcounts_celltypes_orig.ident_GEX3_DIMRED1.csv", row.names = TRUE)

```


## SFig2C Dotplot for cluster marker genes

```{r}

m_genes <- c("MS4A1", "CD79A", "CD79B", "CD3E", "CD3D", "CD3G", 
             "IL7R", "IL32", "FCN1", "LYZ", "CD14","JCHAIN", "MZB1")

p <- DotPlot(gex, 
        features = m_genes, 
        group.by = "seurat_clusters",
        dot.scale = 4
        ) +
  theme_classic() +
  xlab("") + 
  ylab("Seurat KNN clusters") + 
  ggtitle("Cluster marker genes") +
  coord_flip() + 
  theme(axis.text.x = element_text(angle = 90, 
                                       hjust = 1, 
                                       vjust = 0.5,
                                       size = 8),
        panel.grid.major = element_line(colour = "grey", size = rel(0.5)))

ggsave2(paste0("../results/SFig2C.png"), 
        p, 
        width = 12, height = 14, unit = "cm") 

```


## Save rds GEX3_DIMRED1.rds

```{r}

saveRDS(gex, file = "../results/GEX3_DIMRED1.rds")
#gex <- readRDS(file = "../results/GEX3_DIMRED1.rds")

gex@meta.data$SSA[gex@meta.data$orig.ident == "P024"] <- "YES"
gex@meta.data$SSB[gex@meta.data$orig.ident == "P024"] <- "YES"
gex@meta.data$patient_group[gex@meta.data$orig.ident == "P024"] <- "SSAB"
table(gex@meta.data$patient_group, gex@meta.data$orig.ident)

```


## Print sessionInfo()

```{r}

sessionInfo()

```
