# Dimensionality reduction and clustering of B-cells {#Dimred2}

## Load packages

```{r}

source("./00_dependencies.R")

```


## Load data

```{r}

gex <- readRDS("../results/GEX_BCELLS.rds")
annot <- readRDS("../suppl/annot/refdata-gex-GRCh38-2020-A.genes.gtf.annot.rds")

gex

```


## Get variable features B-cells
 
```{r}

gex <- FindVariableFeatures(gex, 
                            selection.method = "vst", 
                            nfeatures = 4000)

p1 <- LabelPoints(plot = VariableFeaturePlot(gex), 
                  points = head(VariableFeatures(gex), 80), 
                  repel = TRUE,
                  xnudge = 0,
                  ynudge = 0,
                  max.overlaps = Inf) + 
        theme_classic() +
        ggtitle(paste0(length(VariableFeatures(gex)), " Variable features - B-cells"))


```


## Subset VariableFeatures(gex) to only include biotype == "protein_coding"

```{r}

VariableFeatures(gex) <-
  VariableFeatures(gex)[VariableFeatures(gex) %in%
                          annot$external_gene_name[annot$gene_biotype == "protein_coding"]]

p2 <- LabelPoints(plot = VariableFeaturePlot(gex), 
                  points = head(VariableFeatures(gex), 80), 
                  repel = TRUE,
                  xnudge = 0,
                  ynudge = 0,
                  max.overlaps = Inf) + 
        theme_classic() +
        ggtitle(paste0(length(VariableFeatures(gex), " Variable features - B-cells (protein_coding)")))


ggsave2("../results/VariableFeatures_B-cells.png",
        plot_grid(p1, p2, ncol = 2, labels = "AUTO", align = "h")
        , width = 30, height = 15, unit = "cm")


```


## Scale data B-cells

```{r}

gex <- ScaleData(gex, 
                 features = VariableFeatures(object = gex),
                 vars.to.regress = c("nFeature_RNA"))


```


## PCA B-cells

```{r}

#RunPCA() takes around 2h for this data set using 16 cores

gex <- RunPCA(gex,
              assay = "RNA",
              features = VariableFeatures(object = gex),
              npcs = 100,
              reduction.name = "pca_2",
              verbose = TRUE)

gex@assays$RNA@scale.data <- matrix(0); gc() #remove scale.data to reduce object size


```


## Run Harmony B-cells

```{r}

gex <- RunHarmony(
  gex,
  group.by.vars = "orig.ident",
  reduction = "pca_2",
  reduction.save = "harmony_2",
  assay = "RNA",
  project.dim = FALSE, #project.dim = FALSE needed for seurat object v4.0.0??
  verbose = TRUE
)


```


## plot PCA and harmony B-cells

```{r}

p1 <- DimPlot(object = gex, 
               reduction = "pca_2", 
               pt.size = .1, 
               group.by = "orig.ident", 
               raster = FALSE) + 
            NoLegend() + 
            ggtitle("PCA B-cells, by Sample")

p2 <- DimPlot(object = gex, 
               reduction = "harmony_2", 
               pt.size = .1, 
               group.by = "orig.ident", 
               raster=FALSE) + 
                NoLegend() + 
                ggtitle("Harmony B-cells, by Sample")

ggsave2("../results/PCA_harmony_B-cells.png", 
        plot_grid(p1, p2, ncol = 2, labels = "AUTO", align = "h"), 
        width = 30, height = 15, unit = "cm")


```


## Run UMAP for B-cells in 2D and 3D

```{r}

#2D
gex <- RunUMAP(
  gex,
  dims = 1:50,
  reduction = "harmony_2",
  metric = "correlation",
  reduction.name = "umap_2",
  min.dist = 0.05, #local cell separation
  spread = .3, #global cell separation
  n.neighbors = 30,
  repulsion.strength = 0.1, #~2x min.dist, repulsion to cells fr annat cluster, global
  negative.sample.rate = 50, #ant ggr n.neighbors celler, global distance
  n.epochs = 100,
  n.components = 2
)

#3D
gex <- RunUMAP(
  gex,
  dims = 1:50,
  reduction = "harmony_2",
  metric = "correlation",
  reduction.name = "umap_2_3d",
  min.dist = 0.05, #local cell separation
  spread = .3, #global cell separation
  n.neighbors = 30,
  repulsion.strength = 0.1, #~2x min.dist, repulsion to cells fr annat cluster, global
  negative.sample.rate = 50, #ant ggr n.neighbors celler, global distance
  n.epochs = 100,
  n.components = 3,
  ret_extra = c("model", "nn", "fgraph")
)


```


## Clustering B-cells

```{r}

gex <- FindNeighbors(gex, 
                     reduction = "harmony_2", 
                     dims = 1:50,
                     k.param = 15,
                     verbose = TRUE)

gex <- FindClusters(gex, 
                    resolution = 1.5, 
                    verbose = TRUE, 
                    graph.name = "RNA_snn")

table(gex@meta.data$seurat_clusters)

pdf("../results/UMAP_B-cells_clusters.pdf", width = 12, height = 12, useDingbats = F)
mypar()
plot_meta(gex, "umap_2", "seurat_clusters", label = T)
dev.off()


```


## hclust 
### test cluster relations for B-cell clusters

```{r}

cluster_means <-
  sapply(as.character(unique(gex$seurat_clusters)), function(x) {
    rowMeans(gex@assays$RNA@data[, gex$seurat_clusters == x])
  })

adj <- (1 - cor(cluster_means)) / 2
h <- hclust(as.dist(adj), method = "ward.D2")

png("../results/hclust_clusters_B-cells_clustRes.png", 
    width = 1000, 
    height = 1000, 
    units = "px",
    pointsize = 20)
par(las=2)
plot(as.dendrogram(h))
dev.off()


```


## Plot 2D UMAP B-cells

```{r}

p1 <- DimPlot(gex, 
                  reduction = "umap_2", 
                  pt.size = .1, 
                  group.by = "orig.ident", 
                  raster=FALSE) + 
              NoLegend() + 
              ggtitle("UMAP B-cells, by Sample")

p2 <- DimPlot(gex, 
                  reduction = "umap_2", 
                  pt.size = .1, 
                  group.by = "seurat_clusters", 
                  label = TRUE, 
                  repel = TRUE, 
                  raster=FALSE) + 
              NoLegend() + 
              ggtitle("UMAP B-cells, by Cluster")

ggsave2("../results/UMAP_B-cells.png", 
        plot_grid(p1, p2, ncol = 2, labels = "AUTO", align = "h"), 
        width = 30, height = 15, unit = "cm")


```


## Plot 3D UMAP B-cells
 
```{r}

df <- data.frame(gex@reductions$umap_2_3d@cell.embeddings)
df <- data.frame(df, seurat_clusters = gex$seurat_clusters, orig.ident = gex$orig.ident)
colnames(df)[1:3] <- c("UMAP_1","UMAP_2","UMAP_3")
pal <- c(scales::hue_pal()(8), 
         RColorBrewer::brewer.pal(9, "Set1"),  
         RColorBrewer::brewer.pal(8, "Set2") )


####------- by cluster
p_State <- plot_ly(df, 
                   x = ~UMAP_1, 
                   y = ~UMAP_2, 
                   z = ~UMAP_3, 
                   color = ~seurat_clusters,
                   colors = pal, size=.5) %>%  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'UMAP_1'), 
                      yaxis = list(title = 'UMAP_2'), 
                      zaxis = list(title = 'UMAP_3')))

htmlwidgets::saveWidget(p_State, "../results/UMAP_B-cells_3d_clusters.html")


####------- by orig.ident
p_State <- plot_ly(df, 
                   x = ~UMAP_1, 
                   y = ~UMAP_2, 
                   z = ~UMAP_3, 
                   color = ~orig.ident,
                   colors = pal, size = .5) %>%  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'UMAP_1'), 
                      yaxis = list(title = 'UMAP_2'), 
                      zaxis = list(title = 'UMAP_3')))

htmlwidgets::saveWidget(p_State, "../results/UMAP_B-cells_3d_orig.ident.html")


```


## Plot marker genes B-cells

```{r}

ig.features <- c("IGHA1", "IGHA2", "IGHG1", "IGHG2", "IGHG3", "IGHG4",  "IGHD", "IGHE", "IGHM")
b.markers <- c("CD79A", "CD79B", "MS4A1", "CD19", "CD27", "IGHA1", "IGHD", "IGHM", "JCHAIN", "MME")
pbmc.markers <- c("MS4A1", "CD19", "CD27", "CD79A", "GNLY", "CD3E", "CD14", "LYZ", "CD8A")
my_pars <- c("nCount_RNA", "nFeature_RNA", "percent_mito", "percent_ribo", "percent_hb")

#pbmc
p.pbmc.markers <- FeaturePlot(gex, 
                                features = pbmc.markers, 
                                reduction = "umap_2", 
                                raster = FALSE)

ggsave2("../results/FeaturePlot_B-cells_pbmcMarkers.png", 
        p.pbmc.markers, 
        width = 25, height = 25, unit = "cm")

####-----  B markers
p.b.markers <- FeaturePlot(gex, 
                             features = b.markers, 
                             reduction = "umap_2", 
                             raster = FALSE)

ggsave2("../results/FeaturePlot_B-cells_BMarkers.png",
        p.b.markers, 
        width = 25, height = 25, unit = "cm")

####-----  ig markers
p.ig.markers <- FeaturePlot(gex, 
                              features = ig.features, 
                              reduction = "umap_2", 
                              raster = FALSE)

ggsave2("../results/FeaturePlot_B-cells_IGMarkers.png",
        p.ig.markers, 
        width = 25, height = 25, unit = "cm")

####-----  my_pars
p.myPars <- FeaturePlot(gex, 
                          features = my_pars , 
                          reduction = "umap_2", 
                          raster = FALSE, 
                          order = TRUE)
ggsave2("../results/FeaturePlot_B-cells_myPars.png", 
        p.myPars, 
        width = 25, height = 25, unit = "cm")


```


